Software architectures to integrate workflow engines in science gateways

### 2016-10-13 answers.tex
Small fixes

  The fact that the patterns may be mixed in real implementations is,
  in our opinion, another motivation for our abstraction effort as it
  helps ``untangling'' the various architectural concepts that may be
  present in a real system. This is particularly true for the abstract
  architectures (nested workflows and workflow conversion) that may be
  instantiated in various ways. Our abstractions highlight that other
  instantiation models may be worth designing, for instance nested
  workflows with task encapsulation, as mentioned in Section 5.2. in
  the Discussion.

@@

\revised{The analysis presented in the paper is intended for
experts in science gateway and workflow engine design. It is an
abstraction effort to identify and evaluate the fundamental
architectural patterns that are encountered while integrating workflow
engines and science gateways. In real systems, such patterns are often
intertwined due to considerations related to the historical and
technical context of software projects. The following work helps
identify the key patterns in systems and highlight potential
improvements.}

@@

\begin{answer}
  S1 denotes the ability of the system to start multiple engine
  instances to execute different workflows, while S2 denotes the
  possibility to execute a single workflow with multiple engines. We
  revised the wording in the description of S1 (page 11) to clarify
  this point. We find the current description of S2 clear enough on
  this point: ``[S2] measures the ability to distribute the execution of
  \emph{a single workflow} among different engine instances.''
\end{answer}

@@

We definitely agree with the reviewer that interaction (g) is
  a tricky one that may lead to critical errors. We added the
  following sentence to the paragraph on interaction (g) in Section
  5.2. of the discussion:

@@

This sentence complements the note already made in section 3.7:
``Note, however,
that implementing interaction g can require sub-
stantial work depending on the complexity of the
workflow language used by the new engine.``
\end{answer}

@@

  As discussed previously, we share the reviewer's concerns
  about the feasibility of workflow conversion (see comments on
  interaction (g) in Section 5.2). The performance of workflow
  conversion is strongly related to the availability of a validated
  interaction (g), which is very dependent on the particular workflow
  engines under consideration. We believe that the issue raised by the
  reviewer basically comes down to the weight given to software
  components and interactions in the analysis. However, using
  different weights for different components or interactions would
  open an infinite number of configurations for discussion. For
  instance, in some cases, the workflow engine already comes with a
  workflow service which therefore doesn't have to be included. To
  address this issue, we shared a spreadsheet at
  \url{https://frama.link/LrPRsCQp} that allows editing the following
  weights:
  \begin{itemize}
  \item Weight given to software components and interactions in
  ``integration'', ``extensibility'' and ``robustness'' criteria.
  \item Weight given to metrics in a criterion (e.g., E$_1$ vs. E$_2$ in
  ``extensibility'').
  \item Weight given to criteria in overall evaluation of Table 3.
\end{itemize}
The paper references this spreadsheet wherever equal weights are
assumed. The readers and reviewers are invited to rely on this
spreadsheet for any issue related to the importance given to a
specific component, interaction, metric or criterion in the
analysis. In the paper, we assumed that all weights equal 1, which we
believe is a reasonable choice although inherently limited.

@@

The reviewer is right that re-purposing software components
  and interactions can lead to revisit the evaluation. We updated the following sentence in section 5.2. to better acknowledge this point:

The particular technical or historical context of a science gateway
project may also influence the choice of an architecture to
integrate workflow engines. For instance, many workflow engines are
already available as web services, which tends to favor service
invocation (\revised{in particular when interactions can be re-purposed}), and other science gateways may have strongly tested task
and data control features (interactions \texttt{b} and \texttt{c}),
which would favor task encapsulation. Similarly, adding a new type of engine
may facilitate integration of a new infrastructure when interactions
\texttt{b} and \texttt{c$_2$} are already available. The migration
cost between architectures has been ignored as well.

@@

\begin{answer}
We updated the definition of R$_1$ and R$_2$ to include
  positive examples as well as negative ones. In addition, the
  specific components included in these metrics are represented in red
  on Figure 2.
\end{answer}

@@

\begin{answer}
\texttt{c}, \texttt{c$_1$} and \texttt{c$_2$} are three different
interactions, which explains the confusion. \texttt{c$_2$} is always
specific to workflow execution because it is defined as the
interaction between the workflow engine and the infrastructure, used
to transfer workflow data (inputs, outputs or intermediate
results). \texttt{c$_1$} is not specific to workflow execution because
it is used by the science gateway to transfer any data regardless of
workflow executions. Letter \texttt{c} was used on the task
encapsulation diagram in Figure 2 to represent that the science gateway transfers
both user and workflow data. We now use \texttt{c$_1$} for all the
data transfers initiated by the science gateway, and \texttt{c$_2$}
for all the data transfers initiated by the workflow engine,
regardless of the type of data transferred. Consequently, we now use
\texttt{c$_1$} instead of \texttt{c} on the task encapsulation diagram in Figure 2.

@@

\begin{answer}
Yes, as for any other metric, high values indicate worse
performance. This is explained at the beginning of section 3:
``Criteria break down into specific metrics where lower value
indicates better performance.''
\end{answer}

@@

\begin{answer}
  We agree with the reviewer that using WSRP (Web Services for Remote
  Portlets) changes the architecture. When portlets are deployed on
  different hosts, we think that the architecture becomes service
  invocation as the portlets then behave as individual software
  components according to our definitions. We added the following comment as a footnote in the presentation of tight integration in section 2.2:
\revised{Note that the
    particular case of portlets can be a bit ambiguous: when portlets
    are deployed remotely, using a protocol such as Web Services for
    Remote Portlets (WSRP), they become independent software
    components according to our definition because they could be
    operated by different teams on different hardware systems. In this
    case, the system should be classified in the service invocation
    architecture since the portlets actually become Web
    Services. The same comment holds for remote object
    invocation.}
\end{answer}

@@

\begin{answer}
  We call ``white-box'' workflow a workflow
  where detailed information about individual tasks is available,
  for instance task statuses, execution logs and resource
  consumption. On the contrary, a ``black-box'' workflow only
  displays global information such as the workflow status and the
  logs of the workflow engine. Fine-grained information about
  workflow tasks helps workflow users (scientists)
  identify errors, it allows gateway administrators to troubleshoot
  executions, and it facilitates debugging by workflow
  developers.

  Multiple software layers affect the accuracy of the information
  reported about the tasks' internals because each layer potentially
  shields or modifies information. Task ids, for instance, may not
  be properly linked to workflow activities if delivered through a
  software layer that hasn't been instrumented for that. \todo{refer
    to olabarriaga2014}.

  The paragraph on fine-grained debugging has been edited accordingly, at the end of Section 3.
\end{answer}

@@

\begin{answer}
  We added a section in the paper (current section 6) to show how our
  evaluation framework could be used in the design of new systems.
\end{answer}

@@

We totally agree with the reviewer that Apache Airavata and
WS-PGrade/gUSE are gateway frameworks. This is how they are presented
in the text. Instead of providing an incomplete list of gateways that
use these frameworks in Table 1, we now refer explicitly to ``gateways
based on Apache Airvata'', ``gateways based on WS-PGRADE/gUSE'' and
``gateways based on the Catania Science Gateway
Framework''. Consistently, the table doesn't mention the individual
gateways based on these frameworks any more.
\end{answer}

@@

\begin{answer}
  The task submission framework is part of the infrastructure
  and its nature doesn't influence the architecture used to integrate
  workflow engines in the science gateway. Formally, it is part of
  interaction \texttt{b} (see definition in 2.1).

@@

  The reviewer's use case, however, suggest that the workflow engine
  is split in two parts: (1) a high-level part integrated in the
  science gateway, which submits ``entire workflows'' to the
  infrastructure; (2) a lower-level part, executed on the
  infrastructure, which executes such ``entire workflows''. It seems
  to us that this matches the definition of a nested workflow model,
  where the parent engine is integrated in the science gateway, and
  the child engine runs on the infrastructure. This is another
  instantiation of the nested workflow abstract model, which is not
  evaluated in the paper as mentioned in section 5.2: "\emph{The
    abstract nested workflows and workflow conversion patterns were
    instantiated with service invocation so that they can be analyzed
    in the same framework as the other architectures. Other types of
    instantiation, for instance nested workflows with task
    encapsulation, could also be envisaged.}". It sounds like the
  architecture described by the reviewer is a nested workflow pattern
  instantiated with a tight integration (parent engine) and task
  encapsulation (child engine).

  We found this scenario interesting and we evaluated it in our
  framework. The analysis is reported in the new section 6 of the
  paper.
\end{answer}

@@

\begin{answer}
  We followed the recommendation of the reviewer and created a new
  Section 3 to present and discuss these examples. The caption of
  Figures 5 and 6 is still long though, because we think that the
  detailed references to the interaction numbers on the Figures should
  be part of them, so that the reader doesn't have to constantly
  navigate between the text and the Figures.
\end{answer}

@@

We find the particular ponderation method suggested by the reviewer
(normalization by the number of components or interaction in the
architecture) a bit unfortunate because it would give different
ponderations to the same component or interaction depending on the
architecture in which it is used, which we think would obfuscate the
comparison.
\end{answer}

@@

\begin{answer}
  While we agree that the Sub-Sections in Section 3 look a bit
  repetitive, we believe that the metrics evaluation reported in Table 2 needs to be
  properly justified. 
\end{answer}

@@

\begin{answer}
  Service invocation and Task encapsulation are fundamentally different for the following reasons:
  \begin{itemize}
  \item In task encapsulation, the workflow engine runs on the infrastructure while in service invocation, it is hosted by a dedicated service.
  \item In task encapsulation, the workflow engine is executed as a task of the infrastructure while in service invocation, it is invoked through a service call. That means, in particular, that in task encapsulation different workflow engines are executed through the same protocol (a task), while different types of service calls might be involved in service invocation (e.g. REST or SOAP Web-Service).
  \item In task encapsulation, the workflow engine does not transfer data to the infrastructure while in service invocation it does (interactioni \texttt{c$_2$}.
  \item In task encapuslation, the science gateway submits tasks to the infrastructure, while in service invocation the workflow engine does.
  \end{itemize}
  Therefore, while we acknowledge that task encapsulation and service
  invocation share characteristics, we believe that their differences
  can be clearly inferred from section 2 of the paper. 
\end{answer}

@@

If a new child engine is added to the nested model then the following interactions an and components have to be updated:
\begin{enumerate}
\item A workflow service has to be created for the workflow engine.
\item The parent engine has to be interfaced with the new workflow
  service, i.e., the workflow service should be able to digest ORE
  bundles, or the parent engine should convert ORE bundles to the
  format supported by the workflow service. ORE bundles will be
  exchanged through interaction \texttt{d}.
 \item The child engine has to be able to transfer data to the
   infrastructure used in the science gateway, i.e. interaction
   \texttt{c$_2$} has to be implemented.
 \item The child engine has to be able to submit tasks to the infrastructure
   used in the science gateway, i.e. interaction \texttt{b} has to be
   implemented,
\end{enumerate}
This is why E$_4$=4 for the nested model.

The situation described by the reviewer focuses on item 1 of the pool
extension and item 1 and 2 of the nested workflow extension
(infrastructure is not a discriminant factor here). First, the
reviewer suggest that the availability of ORE bundles in the SHIWA
framework standardizes the communication with workflow engines so that
it becomes possible to have a generic workflow service that can
consume and produce ORE bundles regardless of the particular engine(s)
it wraps. So indeed, item 1 of the nested case disappears. This is a
typical case where, as we wrote in section 7.2, \emph{The particular
technical or historical context of a science gateway project may also
influence the \revised{evaluation} of an architecture.} In this
case, the possible availability of a generic workflow service based on
ORE bundles is contextual information that can hardly be integrated in
the generic analysis. We believe that our framework helps evaluate
the impact of technologies such as ORE bundles on architectures, as
discussed here.

Second, the
reviewer suggests that the integration of the workflow engine in the
agent (pool model) is easier than in the service because the engine
has to produce and consume ORE bundles in the former case while it
only has to consume them in the latter case. This is true again, but
it comes down to the discussion on the weights of the different
software components. In this particular case, it would make sense to
assign a higher weight to the agent than to the service, which could
be done through the public spreadsheet.  
\end{answer}

@@

\begin{answer}
  We totally agree with the reviewer. However, we think that this
  issue cannot be addressed at the pure architectural level as it
  requires accurate technical information about systems. In some
  systems, the science gateway may become a massive piece of
  unmaintainable code as described by the reviewer, and in some other
  cases, it may have a neat internal architecture that allows for easy
  integration of workflow engines. Once again, we believe that the
  answer to this issue is about assigning weights to the different
  components and interactions, which is now possible through the
  public spreadsheet. For instance, if the science gateway code is
  known to be ``untouchable'' due to a cumbersome, then a high weight
  should be put on the science gateway component. 
\end{answer}

@@

\begin{answer}
We added the following paragraphs to the introduction to section 3:

\revised{Note that the metrics defined hereafter aim at providing a framework
to evaluate architectures independently from any particular system,
regardless of any technical constraint such as the use of a specific
programming language or the internal architecture of the science
gateway. The goal is to abstract architectural discussions from the
particular technical context of a real system. Some degree of
system-specificity can be introduced by assigning weights to
interactions and components through a public spreadsheet available at
\url{https://frama.link/LrPRsCQp}. In particular, sheet "Components
and interactions" contains the weight given to these entities in the
metrics and can be used to produce new versions of
Table 2 that fit the particular technical context
of a real system. For instance, a high weight could be assigned to the
science gateway if its internal architecture is such that it can
hardly be extended in practice.}

\revised{In addition, all the metrics in a criterion contribute
  equally to the global score used for the criterion. Column B of
  sheet ``Metrics'' of the public spreadsheet may be used to assign
  different weights to each metric to adjust the specific context of a
  study.}
\end{answer}

>>>>

  The fact that the patterns may be mixed in real implementations is,
  in our opinion, another motivation for our abstraction effort as it
  helps ``untangling'' the various architectural concepts that may be
  present in a real system. This is particularly true for the abstract
  architectures (nested workflows and workflow conversion) that may be
  instantiated in various ways. Our abstractions highlight that other
  instantiation models may be worth designing, for instance nested
  workflows with task encapsulation, as mentioned in Section 6.2. (Discussion).

@@

\revised{This analysis is intended for experts of science gateway
  and workflow engine design. It is an abstraction effort to identify
  and evaluate the fundamental architectural patterns that are
  encountered while integrating workflow engines and science
  gateways. In real systems, such patterns are often intertwined and
  obfuscated due to the historical and
  technical context of software projects.}

@@

\begin{answer}
  S1 denotes the ability of the system to start multiple engine
  instances to execute different workflows, while S2 denotes the
  possibility to execute a single workflow with multiple engines. We
  revised the wording in the description of S1 (Section 4.4) to clarify
  this point.
\end{answer}

@@

We definitely agree with the reviewer that interaction (g) is
  a tricky one that may lead to critical errors. We added the
  following sentence to the paragraph on interaction (g) in Section
  6.2. of the discussion:

@@

This sentence complements the note already made in section 5.6:
``Note, however, that implementing interaction g can require
substantial work depending on the complexity of the workflow language
used by the new engine.``
\end{answer}

@@

  As discussed previously, we share the reviewer's concerns about the
  feasibility of workflow conversion (see comments on interaction (g)
  in Section 6.2). The performance of workflow conversion is strongly
  related to the availability of a validated interaction (g), which is
  very dependent on the particular workflow engines under
  consideration. We believe that the issue raised by the reviewer
  basically comes down to the weight given to particular software
  components and interactions in the analysis (interaction \texttt{g}
  in this case). However, using different weights for different
  components or interactions would open an infinite number of
  configurations for discussion. For instance, in some cases, the
  workflow engine already comes with a workflow service which
  therefore doesn't have to be counted in the analysis. To address this issue, we
  shared a spreadsheet at \url{https://frama.link/LrPRsCQp} that
  allows editing the following weights:
  \begin{itemize}
  \item Weight given to software components and interactions in
  ``integration'', ``extensibility'' and ``robustness'' criteria.
  \item Weight given to metrics in a criterion (e.g., E$_1$ vs. E$_2$ in
  ``extensibility'').
  \item Weight given to criteria in overall evaluation of Table 3.
\end{itemize}
The paper references this spreadsheet wherever equal weights are
assumed. The readers and reviewers are invited to rely on this
spreadsheet for any issue related to the importance given to a
specific component, interaction, metric or criterion in the
analysis. In the paper, we assumed that all weights equal 1, which we
believe is a reasonable choice although inherently limited.

@@

The reviewer is right that re-purposing software components
  and interactions can lead to revisit the evaluation. We updated the following sentence in section 6.2. to better acknowledge this point:

``The particular technical or historical context of a science gateway
project may also influence the choice of an architecture to
integrate workflow engines. For instance, many workflow engines are
already available as web services, which tends to favor service
invocation (\revised{in particular when interactions can be re-purposed}), and other science gateways may have strongly tested task
and data control features (interactions \texttt{b} and \texttt{c}),
which would favor task encapsulation. Similarly, adding a new type of engine
may facilitate integration of a new infrastructure when interactions
\texttt{b} and \texttt{c$_2$} are already available. The migration
cost between architectures has been ignored as well.''

@@

\begin{answer}
  We updated the definition of R$_1$ and R$_2$ to include positive
  examples in addition to negative ones. In addition, the specific
  components included in these metrics are represented in red on
  Figure 2.
\end{answer}

@@

\begin{answer}
\texttt{c}, \texttt{c$_1$} and \texttt{c$_2$} were three different
interactions, which explains the confusion. \texttt{c$_2$} is always
specific to workflow execution because it is defined as the
interaction between the workflow engine and the infrastructure, used
to transfer workflow data (inputs, outputs or intermediate
results). \texttt{c$_1$} is not specific to workflow execution because
it is used by the science gateway to transfer any data regardless of
workflow executions. Letter \texttt{c} was used on the task
encapsulation diagram in Figure 2 to represent that the science gateway transfers
both user and workflow data. We now use \texttt{c$_1$} for all the
data transfers initiated by the science gateway, and \texttt{c$_2$}
for all the data transfers initiated by the workflow engine,
regardless of the type of data transferred. Consequently, we now use
\texttt{c$_1$} instead of \texttt{c} on the task encapsulation diagram in Figure 2.

@@

\begin{answer}
  Yes, as for any other metric, high values indicate worse
  performance. This is explained at the beginning of section 3:
  ``Criteria break down into specific metrics where \emph{lower value
    indicates better performance}.''
\end{answer}

@@

\begin{answer}
  We agree with the reviewer that using WSRP (Web Services for Remote
  Portlets) changes the architecture. When portlets are deployed on
  different hosts, the architecture becomes service invocation as the
  portlets then behave as individual software components according to
  our definitions. We added the following comment as a footnote in the
  presentation of tight integration in section 2.2: \revised{Note that
    the particular case of portlets can be a bit ambiguous: when
    portlets are deployed remotely, using a protocol such as Web
    Services for Remote Portlets (WSRP), they become independent
    software components according to our definition because they could
    be operated by different teams on different hardware systems. In
    this case, the system should be classified in the service
    invocation architecture since the portlets actually become Web
    Services. The same comment holds for remote object invocation.}
\end{answer}

@@

\begin{answer}
  We call ``white box'' workflow a workflow
  where detailed information about individual tasks is available,
  for instance task statuses, execution logs and resource
  consumption. On the contrary, a ``black box'' workflow only
  displays global information such as the workflow status and the
  logs of the workflow engine. Fine-grained information about
  workflow tasks helps workflow users (scientists)
  identify errors, it allows gateway administrators to troubleshoot
  executions, and it facilitates debugging by workflow
  developers.

  Multiple software layers affect the accuracy of the information
  reported about the tasks' internals because each layer potentially
  shields or modifies information. Task ids, for instance, may not be
  properly linked to workflow activities if delivered through a
  software layer that hasn't been instrumented for that. This issue is
  also mentioned in the following paragraph extracted from [34]
  (reference cited in Section 4.5):
\begin{quote}
  From the workflow system perspective, the operator needs interfaces
  to facilitate DCI configuration, identification and solution of the
  problems encountered during workflow execution. In practice this can
  be challenging because the abstract layer introduced by the workflow
  infrastructure may hamper debugging and troubleshooting, as it may
  hide too much concrete level information.
\end{quote}

  The paragraph on fine-grained debugging has been edited accordingly, at the end of Section 4.5.
\end{answer}

@@

\begin{answer}
  We added a complete new section in the paper (Section 7) to show how our
  evaluation framework could be used in the design of new systems.
\end{answer}

@@

We totally agree with the reviewer that Apache Airavata and
WS-PGrade/gUSE are gateway frameworks. Table 1 now refers explicitly
to ``gateways based on Apache Airvata'', ``gateways based on
WS-PGRADE/gUSE'' and ``gateways based on the Catania Science Gateway
Framework''. 
\end{answer}

@@

\begin{answer}
  The task submission framework (e.g. SAGA) is part of the infrastructure
  and its nature doesn't influence the architecture used to integrate
  workflow engines in the science gateway. Formally, it is part of
  interaction \texttt{b} (see definition in 2.1).

@@

  The reviewer's use case, however, suggest that the workflow engine
  is split in two parts: (1) a high-level part integrated in the
  science gateway, which submits ``entire workflows'' to the
  infrastructure; (2) a lower-level part, executed on the
  infrastructure, which executes such ``entire workflows''. It seems
  to us that this matches the definition of a nested workflow model,
  where the parent engine is integrated in the science gateway, and
  the child engine runs on the infrastructure. This is another
  instantiation of the nested workflow abstract model, which is not
  evaluated in the paper. Several other instantiations of the abstract
  models could be envisaged, as mentioned in section 6.2: "\emph{The
    abstract nested workflows and workflow conversion patterns were
    instantiated with service invocation so that they can be analyzed
    in the same framework as the other architectures. Other types of
    instantiation, for instance nested workflows with task
    encapsulation, could also be envisaged.}".

  The architecture described by the reviewer is a nested workflow
  pattern instantiated with a tight integration (parent engine) and
  task encapsulation (child engine). We found this scenario
  interesting as it illustrates how our framework can be used to help
  designing new systems (although we realize that such a system may
  already exist). We evaluated this hypothetical architecture in our
  framework: the analysis is reported in the new section 7 of the
  paper.
\end{answer}

@@

@@

We find the particular ponderation method suggested by the reviewer
(normalization by the number of components or interaction in the
architecture) a bit unfortunate because it would give different
values to the same component or interaction depending on the
architecture in which it is used, which we think would obfuscate the
comparison.
\end{answer}

@@

\begin{answer}
  While we agree that the Sub-Sections in Section 5 (former Section 3) look a bit
  repetitive, we believe that the metrics evaluation reported in Table 2 needs to be
  properly justified. We contracted the text wherever possible.
\end{answer}

@@

\begin{answer}
  Service invocation and Task encapsulation are fundamentally different for the following reasons:
  \begin{itemize}
  \item In task encapsulation, the workflow engine runs on the infrastructure while in service invocation, it is hosted by a dedicated service.
  \item In task encapsulation, the workflow engine is executed as a task of the infrastructure while in service invocation, it is invoked through a service call. That means, in particular, that in task encapsulation different workflow engines are executed through the same protocol (a task), while different types of service calls might be involved in service invocation (e.g. REST or SOAP Web-Service).
  \item In task encapsulation, the workflow engine does not transfer data to the infrastructure while in service invocation it does (interaction \texttt{c$_2$}.
  \item In task encapuslation, the science gateway submits tasks to the infrastructure, while in service invocation the workflow engine does.
  \end{itemize}
\end{answer}

@@

If a new child engine is added to the nested model then the following interactions an and components have to be updated (a similar analysis can be done when a parent engine is added):
\begin{enumerate}
\item A workflow service has to be created for the workflow engine.
\item The parent engine has to be interfaced with the new workflow
  service, i.e., the workflow service should be able to digest ORE
  bundles, or the parent engine should convert ORE bundles to the
  format supported by the workflow service. ORE bundles will be
  exchanged through interaction \texttt{d}.
 \item The child engine has to be able to transfer data to the
   infrastructure used in the science gateway, i.e. interaction
   \texttt{c$_2$} has to be implemented.
 \item The child engine has to be able to submit tasks to the infrastructure
   used in the science gateway, i.e. interaction \texttt{b} has to be
   implemented,
\end{enumerate}
This is why E$_4$=4 for the nested model when a child engine is added.

The situation described by the reviewer focuses on item 1 of the pool
extension and item 1 and 2 of the nested workflow extension
(infrastructure is not a discriminant factor here). First, the
reviewer suggests that the availability of ORE bundles in the SHIWA
framework standardizes the communication with workflow engines so that
it becomes possible to have a generic workflow service that can
consume and produce ORE bundles regardless of the particular engine(s)
it wraps. So indeed, item 1 of the nested case disappears. This is a
typical case where, as we wrote in section 7.2, \emph{The particular
technical or historical context of a science gateway project may also
influence the \revised{evaluation} of an architecture.} In this
case, the possible availability of a generic workflow service based on
ORE bundles is contextual information that can hardly be integrated in
the generic analysis. However, our framework helps evaluate
the impact of technologies such as ORE bundles on architectures, as
discussed here.

Second, the reviewer suggests that the integration of the workflow
engine in the agent (pool model) is easier than in the service because
the engine has to produce and consume ORE bundles in the former case
while it only has to consume them in the latter case. This is true
again, but we think that it comes down to a discussion on the weights
of the different software components. In this particular case, it
would make sense to assign a higher weight to the agent than to the
service, because integrating a workflow engine in the agents requires
more work than integrating it in a service.

To address this issue, we
  shared a spreadsheet at \url{https://frama.link/LrPRsCQp} that
  allows editing the following weights (see also answers to Reviewers \#1 and \#2):
  \begin{itemize}
  \item Weight given to software components and interactions in
  ``integration'', ``extensibility'' and ``robustness'' criteria.
  \item Weight given to metrics in a criterion (e.g., E$_1$ vs. E$_2$ in
  ``extensibility'').
  \item Weight given to criteria in overall evaluation of Table 3.
\end{itemize}
The paper references this spreadsheet wherever equal weights are
assumed. The readers and reviewers are invited to rely on this
spreadsheet for any issue related to the importance given to a
specific component, interaction, metric or criterion in the
analysis. In the paper, we assumed that all weights equal 1, which we
believe is a reasonable choice although inherently limited.
\end{answer}

@@

\begin{answer}
  We totally agree with the reviewer. However, we think that this
  issue cannot be addressed at the pure architectural level as it
  requires accurate technical information about systems. In some
  systems, the science gateway may become a massive piece of
  unmaintainable code as described by the reviewer, and in some other
  cases, it may have a neat internal architecture that allows for easy
  integration of workflow engines. Once again, we believe that the
  answer to this issue is about assigning weights to the different
  components and interactions, which is now possible through the
  public spreadsheet. For instance, if the science gateway code is
  known to be ``untouchable'' due to a bulky architecture, then a high weight
  should be put on the science gateway component. 
\end{answer}

@@

\begin{answer}
We added the following paragraphs to the introduction to Section 4:

\revised{The metrics defined hereafter aim at providing a framework
  to evaluate architectures independently from any particular system,
  regardless of any technical constraint such as the use of a specific
  programming language or the internal architecture of the science
  gateway. The goal is to abstract architectural discussions from the
  particular technical context of a real system. Some degree of
  system-specificity can be introduced in the evaluation by assigning
  weights to interactions and components through a public spreadsheet
  available at \url{https://frama.link/LrPRsCQp}\footnote{The
    spreadsheet may need to be downloaded to be edited.}. In
  particular, sheet ``Components and interactions'' in this
  spreadsheet contains the weight assigned to components and
  interactions, which can be used to produce new versions of
  Table 2 that fit the particular technical
  context of a real system. For instance, a high weight could be
  assigned to the science gateway if its internal architecture is such
  that it can hardly be extended in practice.

  In addition, all the metrics in a criterion contribute
  equally to the global score used for the criterion. Column B of
  sheet ``Metrics'' of the public spreadsheet may be used to assign
  different weights to each metric to adjust the specific context of a
  study.
}
\end{answer}

### 2016-10-13 paper.tex
Small fixes

\correction{This analysis is intended for experts in science gateway
  and workflow engine design. It is an abstraction effort to identify
  and evaluate the fundamental architectural patterns that are
  encountered while integrating workflow engines and science
  gateways. In real systems, such patterns are often intertwined and
  obfuscated due to the historical and
  technical context of software projects.}

@@

Architectures to integrate workflow engines in science gateways are
presented in Figure~\ref{fig:architectures} using the graphical
notations defined in Figure~\ref{fig:notations}. \correction{Software
  components are defined at a granularity such that two components may
  potentially be operated by different teams on different hardware
  systems. For instance, the science gateway and infrastructure are
  usually distinct components. In real systems, software components
  may exist at a finer granularity, to implement functions such as
  authentication, data persistency or logging. Such fine-grained
  components are not covered by our analysis since we focus on the
  interactions between the science gateway, the infrastructure and the
  workflow engine.}  Table~\ref{table:system-classification}
summarizes the classification of a few systems by architecture.

@@

  can be used to homogeneize these interfaces.

@@

See Figure~\subref*{archi:sub-task}. The workflow engine is wrapped in
a particular task that can submit sub-tasks to the science gateway
through interaction \texttt{e}. The workflow engine keeps track of the
dependencies between the sub-tasks, but their execution is delegated
to the science gateway that executes them on the infrastructure
through interaction \texttt{b}. Although the science gateway has no
global vision of the workflow, it completely \correction{controls} the
sub-tasks. \correction{For instance, it may cancel} them when the
\correction{workflow} is canceled. The science gateway may also
implement mechanisms to facilitate the handling of task dependencies,
for instance \correction{lists of tasks that must be completed before a
  particular task can be executed}.

@@

\item Fine-grained debugging (\texttt{\correction{F}$_2$}): availability of
  fine-grained debugging information about workflow tasks (white-box
  workflow). \correction{We call ``white-box'' workflow a workflow
    where detailed information about individual tasks is available,
    for instance task statuses, execution logs and resource
    consumption. On the contrary, a ``black-box'' workflow only
    displays global information such as the workflow status and the
    logs of the workflow engine.} Fine-grained information about
  workflow tasks is required to properly troubleshoot workflow
  executions. \correction{It helps workflow users (scientists)
    identify errors, it allows gateway administrators to troubleshoot
    executions, and it facilitates debugging by workflow
    developers}. \correction{The availability and accuracy of
    fine-grained debugging information} usually depends on the number
  of software layers between the science gateway and the workflow
  engine~\cite{olabarriaga2014}. \texttt{\correction{F}$_2$}=0 means that the
  information is directly accessible in the science gateway,
  \texttt{\correction{F}$_2$}=1 means that it is obtained through one or more
  software interactions.
\end{itemize}
\texttt{\correction{F}$_1$} and \texttt{\correction{F}$_2$} are summed to obtain a total number of missing 
features in the architecture.

@@

The particular case of interaction \texttt{g} (workflow language
conversion) requires particular attention when implementing a real
system since this interaction may not be easily generalizable to any
workflow language. For instance, converting FSL, PSOM or Nipype
pipelines to any other workflow language is problematic because these
engines rely on general-purpose programming languages such as Bash,
Octave/Matlab and Python, which are much richer than scientific
workflow languages. \correction{If not properly validated, workflow
  language conversion could introduce critical errors impacting the
  robustness and correctness of the execution. }

>>>>

\correction{This analysis is intended for experts of science gateway
  and workflow engine design. It is an abstraction effort to identify
  and evaluate the fundamental architectural patterns that are
  encountered while integrating workflow engines and science
  gateways. In real systems, such patterns are often intertwined and
  obfuscated due to the historical and
  technical context of software projects.}

@@

Architectures to integrate workflow engines in science gateways are
presented in Figure~\ref{fig:architectures} using the graphical
notations defined in Figure~\ref{fig:notations}. \correction{Software
  components are defined at a granularity such that two components may
  potentially be operated by different teams on different hardware
  systems. For instance, the science gateway and infrastructure are
  usually distinct components. In real systems, software components
  may exist at a finer granularity, to implement functions such as
  authentication, data persistence or logging. Such fine-grained
  components are not covered by our analysis since we focus on the
  interactions between the science gateway, the infrastructure and the
  workflow engine.}  Table~\ref{table:system-classification}
summarizes the classification of a few systems by architecture.

@@

  can be used to homogenize these interfaces.

@@

See Figure~\subref*{archi:sub-task}. The workflow engine is wrapped in
a particular task that can submit sub-tasks to the science gateway
through interaction \texttt{e}. The workflow engine keeps track of the
dependencies between the sub-tasks, but their execution is delegated
to the science gateway that executes them on the infrastructure
through interaction \texttt{b}. Although the science gateway has no
global vision of the workflow, it \correction{submits the sub-tasks to the
infrastructure} and completely \correction{controls them}. \correction{For instance, it may cancel} them when the
\correction{workflow} is canceled. The science gateway may also
implement mechanisms to facilitate the handling of task dependencies,
for instance \correction{lists of tasks that must be completed before
  a particular task can be executed}.

@@

\item Fine-grained debugging (\texttt{\correction{F}$_2$}): availability of
  fine-grained debugging information about workflow tasks (white box
  workflow). \correction{We call ``white box'' workflow a workflow
    where detailed information about individual tasks is available,
    for instance task statuses, execution logs and resource
    consumption. On the contrary, a ``black box'' workflow only
    displays global information such as the workflow status and the
    logs of the workflow engine.} Fine-grained information about
  workflow tasks is required to properly troubleshoot workflow
  executions. \correction{It helps workflow users (scientists)
    identify errors, it allows gateway administrators to troubleshoot
    executions, and it facilitates debugging by workflow
    developers}. \correction{The availability and accuracy of
    fine-grained debugging information} usually depends on the number
  of software layers between the science gateway and the workflow
  engine~\cite{olabarriaga2014}. \texttt{\correction{F}$_2$}=0 means that the
  information is directly accessible in the science gateway,
  \texttt{\correction{F}$_2$}=1 means that it is obtained through one or more
  software interactions.
\end{itemize}
\texttt{\correction{F}$_1$} and \texttt{\correction{F}$_2$} are summed to obtain a total number of missing 
features in the architecture.

@@

The particular case of interaction \texttt{g} (workflow language
conversion) requires particular attention when implementing a real
system since this interaction may not be easily generalized to any
workflow language. For instance, converting FSL, PSOM or Nipype
pipelines to any other workflow language is problematic because these
engines rely on general-purpose programming languages such as Bash,
Octave/Matlab and Python, which are much richer than scientific
workflow languages. \correction{If not properly validated, workflow
  language conversion could introduce critical errors impacting the
  robustness and correctness of the execution. }
